Fail Faster

### 2025-03-16 paper/paper.tex
incorporate harry comment

The RNG sits at the heart of the hot path.
Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
and generator combinators like \texttt{list} call those basic generators.
\hg{To
generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

@@

\hg{Alternative framing: ``Of course, simply arguing that the RNG is on the hot
path does not guarantee that a faster RNG leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the RNG for \bq\ with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}

In the context of this paper, however, clearly demonstrating the benefits of any of the
above faster options is tricky. Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally\hg{Is this how we're explaining this? I'm not sure that's quite right} identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property. In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.

% \jwc{There are a few aspects of PBT that make it interesting when it comes to PBT choice.
% \begin{itemize}
%   \item Doesn't need to be cryptographic.
%   \item doesn't even \emph{necessarily} need to pass sophisticated staistical
%   tests.  \cn
% \end{itemize}
% }

% \jwc{
%   \begin{itemize}
%     \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
%     slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
%     just how much you can improve bug-finding performance simply by speeding up the RNG.
%   \end{itemize}
% }

% In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
% performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
% We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
% to the same effect without leaving OCaml.

% It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
% the abstraction overhead of the generator DSL itself, but simply due to language
% implementation and algorithm choices. We include this discussion --- and
% the corresponding evaluation
% --- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
% should be engineered for performance.

% \jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}

>>>>

The RNG sits at the heart of the hot path.
Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
and generator combinators like \texttt{list} call those basic generators.
For instance, to generate a single list with 3 elements, 
\hg{To generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

@@

\hg{Alternative framing: ``'': \dots''}
Of course, simply arguing that the RNG is on the hot
path does not guarantee that a faster RNG leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the RNG for \bq\ with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment:
OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.
